Predicting Fire Risk: \\ An Evaluation of the Fire Danger Index

### 2016-01-22 paper.tex
Fixed tables

\usepackage{fullpage,booktabs,amsmath}
\usepackage{gensymb}
\usepackage{natbib}
\usepackage{pdfpages}
\usepackage{listings}

@@

There is limited literature on the effectiveness of the Australian fire danger rating system using statistical techniques, the majority of Australian research uses case studies. Internationally research on fire danger rating systems use standard econometric techniques such as logistic regression or heckman sample selection models. Machine learning and non-parametric estimation are not used extensively in this area.

@@

\begin{table}[!ht]
	\centering
	\begin{tabular}{ll}
		\toprule
		Fire Danger Rating & Forest Fire Danger Index \\
		\midrule
		Low-Moderate              & 0--11 \\
		High                      & 12--24 \\
		Very High                 & 25--49\\
		Severe                    & 50--74\\
		Extreme                   & 75--99 \\
		Code Red                  & 100+ \\
		\bottomrule
	\end{tabular}
	\caption{The Forest Fire Danger Index and the Fire Danger Rating}
	\label{table:FDR}
\end{table}

@@

The CFA divides the state into different districts for management purposes. I use these districts to capture different terrain types, fuel types and population density in Victoria. Dividing the data into these districts allows us to observe any difference in the behaviour of the variables across different parts of the state. Figure \ref{fig:dist} shows a map of the districts and their labels.

\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/cfadis.pdf} 
	\caption{Country Fire Authority Districts} 
	\label{fig:dist} 
\end{figure}

@@

%Graphs, need to redo better
\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/silvan.pdf} 
	\caption{Time series of meteorological variables for Silvan (-37.82, 145.42)} 
	\label{fig:silv} 
\end{figure}

Figure \ref{fig:silv} shows the time series of the meteorological variables for one point in Victoria. The temperature variable is the daily maximum, values are given in Kelvin (\degree C = K - 273.15), and we clearly observe the seasonal cycles with higher temperatures in the summer months and lower temperatures in the winter months. Maximum daily temperatures across the state range from around 2 \degree C to 46\degree  C.

@@

%WS, Temp and FDR contours, see if can change colours to match with the diagram
\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/tw_fdr.pdf} 
	\caption{Wind Speed and temperature coloured by fire danger rating} 
	\label{fig:tw_fdr} 
\end{figure}

The FFDI is most sensitive to temperature and windspeed (\cite{dowdy10}). Figure \ref{fig:tw_fdr} shows two-dimensional density plots for windspeed and temperature coloured by the fire danger rating (only observations from October-March included) and divided by CFA district. The relative sizes of each contour show the small proportion of extreme and code red days and the difference in weather conditions observed in different parts of Victoria. For instance the only districts observing code red days are districts 16, 17 and 18.

@@

% level plot of GS curing, probably needs resizing
\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/cur_lp.pdf} 
	\caption{Level Plot of Grassland Curing } 
	\label{fig:cur_lp} 
\end{figure}

Figure \ref{fig:cur_lp} shows a level plot produced using the rasterVis package \citep{rastervis}. We observe grassland curing values across the state for the week beginning 25th January 2009. The horizontal margin shows the mean curing value for each longitude and the vertical margin shows the mean curing for each latitude. The plot clearly shows the difference in curing values across the state, in particular the difference between the east and west of the state. This is consistent with time series graphs of curing.

@@

% level plot of vapour pressure, level plot may be unnesc could just have std plot
\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/vap_ras.pdf} 
	\caption{Raster plot of vapour pressure 9am } 
	\label{fig:vap_ras} 
\end{figure}

Figure \ref{fig:vap_ras} shows a level plot of the vapour pressure at 9am \citep{rastervis}. The pixel values of the vapour pressure across the state on the 7th of February 2009. We see very clearly in the north-west of the state vapour pressure at 9am is much lower relative to the south-east.

@@

\begin{table}[!ht]
	\centering
	\begin{tabular}{lrrrrrrrr}
		\toprule
		       & ffdi  & temp  & df    & rh    & ws    & curing & rain  & vap \\
		\midrule
		ffdi   & 1.00  & 0.77  & 0.62  & -0.77 & 0.04  & 0.51   & -0.20 & 0.01 \\
		temp   & 0.77  & 1.00  & 0.43  & -0.70 & -0.18 & 0.46   & -0.10 & 0.39 \\
		df     & 0.62  & 0.43  & 1.00  & -0.58 & -0.09 & 0.47   & -0.37 & -0.08 \\
		rh     & -0.77 & -0.70 & -0.58 & 1.00  & 0.19  & -0.49  & 0.31  & 0.06 \\
		ws     & 0.04  & -0.18 & -0.09 & 0.19  & 1.00  & 0.01   & 0.07  & -0.16 \\
		curing & 0.51  & 0.46  & 0.47  & -0.49 & 0.01  & 1.00   & -0.11 & 0.05 \\
		rain   & -0.20 & -0.10 & -0.37 & 0.31  & 0.07  & -0.11  & 1.00  & 0.17 \\
		vap    & 0.01  & 0.39  & -0.08 & 0.06  & -0.16 & 0.05   & 0.17  & 1.00 \\
		\bottomrule
	\end{tabular}
	\caption{Correlation matrix of variables}
	\label{table:correl}
\end{table}

Table \ref{table:correl} shows the cross correlation matrix for the variables used to predict probability of ignition. High correlation can pose a problem in regression models and to a lesser extent machine learning methods. Variables of concern are temperature, drought factor, relative humidity and FFDI and  which have relatively strong correlation coefficients.

@@

%Commented out until final printing, 
%Map of fire ignitions in Victoria
\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/data/firemap.pdf} 
	\caption{Map of fire ignition 2000 - 2009} 
	\label{fig:firemap} 
\end{figure}

%timeseries histogram with count of ignitions per fortnight
\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/data/occurence.pdf} 
	\caption{Number of ignitions by fortnight} 
	\label{fig:occ} 
\end{figure}

@@

Figure \ref{fig:firemap} shows the locations of the fires in the dataset on the map of Victoria. The most noticeable thing in Figure \ref{fig:firemap} is the sparsity of fires in the north-east and north west of the state and around Melbourne. This is due to the division of fire fighting organisations in Victoria. The sparse areas correspond to National Parks where Parks Victoria or the Department of Environment, Land, Water and Planning (DELWP) coordinate fire fighting activities, so many of the fires occurring in these areas are not included in this dataset.

\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/tw_fire.pdf} 
	\caption{Wind Speed and temperature coloured by fire/no fire} 
	\label{fig:tw_fire} 
\end{figure}

Figure \ref{fig:tw_fire} is constructed in the same manner as Figure \ref{fig:tw_fdr} but instead of grouping by the fire danger rating, I group by Fire/No Fire. Windspeed and temperature play the largest role in determining the FFDI yet there is no discernible difference between the distributions of the two variables when there is a fire compared to no fires. This is the first indicator that the FFDI alone may not be a good predictor of probability of ignition.

\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/dh_fire.pdf} 
	\caption{Drought factor and relative humidity coloured by fire/no fire} 
	\label{fig:dh_fire} 
\end{figure}

Plotting the 2D densities of the drought factor and relative humidity in Figure \ref{fig:dh_fire}, ranges of relative humidity don't appear to vary on fire days compared to not fire days but in districts 4, 5, 6, 7, 8, 9, 10, 11, 13, 14 and 24 there is a clear difference in the densities of fire compared to no fire. Small drought factors are mostly associated with non-fire days. In districts 2, 12, 15, 16, 17, 18, 20 and 22 do not have large differences in the two densities, these districts generally have high drought factor and low relative humidity.

\begin{figure}[!ht]
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/gv_fire.pdf} 
	\caption{Grassland curing and vapour pressure coloured by fire/no fire} 
	\label{fig:gv_fire} 
\end{figure}

In Figure \ref{fig:gv_fire} shows the 2D densities for grassland curing and vapour pressure for the fire and non-fire days. Once again we see a clear difference in behaviour across districts.

@@

From Figure \ref{fig:occ} it is clear the majority of bushfires occur during the fire season. I reduce the size of the dataset, taking only observations in Victorian fire season (October to March) for a number of reasons. First, from a policy perspective the FFDI, fire danger rating systems, grassland curing are only calculated during the fire season so using these variables outside the time periods they are available does not make intuitive sense. From a practical perspective, this paper is looking at predicting the types bushfires that threaten lives and property which we know to occur in the summer months.

@@

This final dataset is divided into a testing (30\%) and training set (70\%) randomly sampling across 'Fire' and 'No Fire' such that both datasets have the same proportion of 'Fire' (approximately 0.05).

@@

As a result the logistic model becomes
\begin{equation}
\label{eq logits}
\Pr(Y=1|X)=\frac{\exp(\sum_{k=1}^K f_k(x_k))}{1+\exp(\sum_{k=1}^K f_k(x_k)}
\end{equation}

where $Y$ denotes fire occurrence, $x_k$ denotes the $k^{th}$ predictor and each $f_k$ represents a penalized regression spline.

@@

I address the class imbalance problem in two ways, first in building the model I create balanced samples by under sampling the `no fire' class in a similar manner to \citep{padilla11}.  The second approach is to assess each model's performance in predicting the test data using 'balanced accuracy' rather than overall accuracy \citep{mosley13, japkowicz00}.

Under-sampling the no fire class addresses the problem of imbalanced classes but I also use it to reduce the serial and spatial correlation that is present in the dataset by using a ~near-maximin' sampling method thereby ensuring all the data points in the sample are separate from each from each other in space and time to reduce the impact of correlation.

@@

An example confusion matrix is given in Table~\ref{table:ex}. 
%eg confusion matrix with ABCD
\begin{table}[!ht]
	\centering
	\begin{tabular}{lr|rr|r|}
    & 	\multicolumn{1}{l}{} & \multicolumn{2}{c}{\emph{Model Prediction}} \\
    & 	\multicolumn{1}{l}{} & \multicolumn{1}{r}{No Fire} & \multicolumn{1}{r}{Fire} \\
		\cline{3-5}
\emph{Actual} & No Fire & a   & a   & a+b \\
                 & 		Fire & c   & d   & c + d \\
		\cline{3-5}
                 & 		     & a+c & b+d & N \\
		\cline{3-5}
	\end{tabular}
	\caption{Example confusion matrix}
	\label{table:ex}
\end{table}

Accuracy is defined as:
\begin{equation}
\label{eq acc}
\mbox{Accuracy} = \frac{a + d}{a+b+c+d}
\end{equation}

I also report sensitivity (correctly classified fires dividied by total number of actual fires) and specificity (correctly identified non-fires divided by total number of non-fires), balanced accuracy and kappa:
\begin{align}
\label{eq sens}
\mbox{Sensitivity} &= \frac{a }{a+b} \\
\label{eq spec}
\mbox{Specificity} &= \frac{d}{c+d} \\
\label{eq BA}
\mbox{Balanced Accuracy} &= \frac{\frac{a }{a+b} + \frac{d}{c+d} }{2} \\
\label{eq:K}
\kappa &= \frac{P_0 - P_c}{1-P_c}
\end{align}
where $P_0$ is the total agreement probability and $P_c$ is the Agreement probability which is due to chance (see \cite{kappa} for definitions).

@@

Logistic regression output for the first three models is given in table \ref{table:coefficients}.

@@

I add splines to Model 3 in stages using the confusion matrices for 10-fold cross validation to assess accuracy (using the caret package by \cite{caret}). The best performing model based on balanced accuracy and number of false negatives (incorrectly predicted ~No Fires') is reported in table \ref{table:logf}.

@@

Figure \ref{fig:gam_pl} shows the spline functions for each of the inputs. In logistic models the spline functions are not the marginal effects of the variables on probability of ignition but the marginal effect on the log-odds ratio. Equation \ref{eq logits} can be rearranged to give equation \ref{eq:logodds}, the log-odds ratio \citep{james13}.

\begin{equation}
\label{eq:logodds}
\log(\frac{p(X)}{1-p(X)}) = \sum_{k=1}^{K} f_k (x_k)
\end{equation}

The spline function fitted on the FFDI although non-linear, according to the standard error is not statistically different from zero (also seen in the p-values in table \ref{table:logf}). After taking into account the other variables in the model, FFDI does not appear to have an impact on the log-odds ratio. This may be because the other variables in the model take into account any information contained in the FFDI.

@@

%Output from best performing splines model (used xtable rather than texreg, too inconsistent?)
\begin{table}[!ht]
	\centering
	\begin{tabular}{rrrrr}
		\toprule
		& Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
		\midrule
		(Intercept) & -3.1484 & 0.2757 & -11.4210 & 0.0000 \\ 
		ns(ffdi, 5)1 & -0.2905 & 0.3613 & -0.8043 & 0.4213 \\ 
		ns(ffdi, 5)2 & -0.2304 & 0.4625 & -0.4982 & 0.6183 \\ 
		ns(ffdi, 5)3 & 0.4040 & 0.5589 & 0.7229 & 0.4697 \\ 
		ns(ffdi, 5)4 & -0.4340 & 0.9674 & -0.4487 & 0.6537 \\ 
		ns(ffdi, 5)5 & -0.8815 & 0.7271 & -1.2124 & 0.2254 \\ 
		ns(temp, 1) & 1.8543 & 0.3416 & 5.4276 & 0.0000 \\ 
		ns(rh, 2)1 & 0.9325 & 0.5687 & 1.6397 & 0.1011 \\ 
		ns(rh, 2)2 & -0.6173 & 0.3607 & -1.7115 & 0.0870 \\ 
		ws & -0.0169 & 0.0146 & -1.1541 & 0.2485 \\ 
		ns(df, 5)1 & 1.1093 & 0.2798 & 3.9639 & 0.0001 \\ 
		ns(df, 5)2 & 1.0156 & 0.3115 & 3.2601 & 0.0011 \\ 
		ns(df, 5)3 & 0.7113 & 0.2604 & 2.7319 & 0.0063 \\ 
		ns(df, 5)4 & 0.8071 & 0.5479 & 1.4730 & 0.1408 \\ 
		ns(df, 5)5 & -0.1512 & 0.1950 & -0.7752 & 0.4382 \\ 
		ns(curing, 6)1 & 1.5648 & 0.1388 & 11.2754 & 0.0000 \\ 
		ns(curing, 6)2 & 2.3986 & 0.1874 & 12.7997 & 0.0000 \\ 
		ns(curing, 6)3 & 1.6637 & 0.1640 & 10.1436 & 0.0000 \\ 
		ns(curing, 6)4 & 1.6949 & 0.1335 & 12.6921 & 0.0000 \\ 
		ns(curing, 6)5 & 1.4634 & 0.3436 & 4.2588 & 0.0000 \\ 
		ns(curing, 6)6 & -0.4909 & 0.0817 & -6.0066 & 0.0000 \\ 
		rain & -0.0286 & 0.0056 & -5.0703 & 0.0000 \\ 
		vap & -0.0003 & 0.0067 & -0.0504 & 0.9598 \\ 
		\bottomrule
	\end{tabular}
	\caption{Regression output for Model 4}
	\label{table:logf}
\end{table}

%GAM plots, not sure if this is exactly splines... 
\begin{figure}[!ht]
\null\vfill\noindent
%\includegraphics[page=1,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=2,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
%\includegraphics[page=3,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=4,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
%\includegraphics[page=5,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=6,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
\null
	\caption{Plots of smoothing functions in Model 4 with 95\% point-wise confident interval} 
	\label{fig:gam_pl} 
\end{figure}

The focus of this project is to build an accurate predictor of fire ignition, so out of sample prediction is a more important evaluation tool than hypothesis testing. Table \ref{table:cm2} contains confusion matrices (values are proportions of total number of observations in the test set) based on test data predictions, using a cutoff value of 0.5. The confusion matrix is accompanied by measures of accuracy in table \ref{table:acc2}.

%Confusion matrix cutoff=0.5
\begin{table}[!ht]
	\centering
	\begin{tabular}{r|rr|rr|rr|rr}
		\toprule
		& Model 1 &  & Model 2 &  & Model 3 &  & Model 4 & \\ 
		\midrule
		& Actual   &  &  &  &  & & & \\ 
		Predict & No Fire & Fire & No Fire & Fire & No Fire & Fire & No Fire & Fire\\ 
		\midrule
No Fire & 0.5959 & 0.0030 & 0.4457 & 0.0019 & 0.5111 & 0.0021 & 0.5807 & 0.0016 \\ 
Fire & 0.3989 & 0.0022 & 0.5490 & 0.0034 & 0.4838 & 0.0030 & 0.4142 & 0.0035 \\ 
		\bottomrule
	\end{tabular}
	\caption{Confusion Matrices for Models 1, 2 \& 3 cutoff value 0.5}
	\label{table:cm2}
\end{table}

>>>>

\usepackage{fullpage,booktabs,amsmath}
\usepackage[section]{placeins}
\usepackage{gensymb,microtype,todonotes}
\usepackage{natbib}
\usepackage{pdfpages}
\usepackage{listings}

@@

There is limited literature on the effectiveness of the Australian fire danger rating system using statistical techniques, the majority of Australian research uses case studies. Internationally research on fire danger rating systems use standard econometric techniques such as logistic regression or Heckman sample selection models. Machine learning and non-parametric estimation are not used extensively in this area.

@@

\begin{table}
	\centering
	\begin{tabular}{ll}
		\toprule
		Fire Danger Rating & Forest Fire Danger Index \\
		\midrule
		Low-Moderate              & 0--11 \\
		High                      & 12--24 \\
		Very High                 & 25--49\\
		Severe                    & 50--74\\
		Extreme                   & 75--99 \\
		Code Red                  & 100+ \\
		\bottomrule
	\end{tabular}
	\caption{The Forest Fire Danger Index and the Fire Danger Rating}
	\label{table:FDR}
\end{table}

@@

The CFA divides the state into different districts for management purposes. I use these districts to capture different terrain types, fuel types and population density in Victoria. Dividing the data into these districts allows us to observe any difference in the behaviour of the variables across different parts of the state. Figure~\ref{fig:dist} shows a map of the districts and their labels.

\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/cfadis.pdf} 
	\caption{Country Fire Authority Districts} 
	\label{fig:dist} 
\end{figure}

@@

%Graphs, need to redo better
\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/silvan.pdf} 
	\caption{Time series of meteorological variables for Silvan (-37.82, 145.42)} 
	\label{fig:silv} 
\end{figure}

Figure~\ref{fig:silv} shows the time series of the meteorological variables for one point in Victoria. The temperature variable is the daily maximum, values are given in Kelvin (\degree C = K - 273.15), and we clearly observe the seasonal cycles with higher temperatures in the summer months and lower temperatures in the winter months. Maximum daily temperatures across the state range from around 2 \degree C to 46\degree  C.

@@

%WS, Temp and FDR contours, see if can change colours to match with the diagram
\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/tw_fdr.pdf} 
	\caption{Wind Speed and temperature coloured by fire danger rating} 
	\label{fig:tw_fdr} 
\end{figure}

The FFDI is most sensitive to temperature and windspeed (\cite{dowdy10}). Figure~\ref{fig:tw_fdr} shows two-dimensional density plots for windspeed and temperature coloured by the fire danger rating (only observations from October--March included) and divided by CFA district. The relative sizes of each contour show the small proportion of extreme and code red days and the difference in weather conditions observed in different parts of Victoria. For instance the only districts observing code red days are districts 16, 17 and 18.

@@

% level plot of GS curing, probably needs resizing
\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/cur_lp.pdf} 
	\caption{Level Plot of Grassland Curing } 
	\label{fig:cur_lp} 
\end{figure}

Figure~\ref{fig:cur_lp} shows a level plot produced using the rasterVis package \citep{rastervis}. We observe grassland curing values across the state for the week beginning 25th January 2009. The horizontal margin shows the mean curing value for each longitude and the vertical margin shows the mean curing for each latitude. The plot clearly shows the difference in curing values across the state, in particular the difference between the east and west of the state. This is consistent with time series graphs of curing.

@@

% level plot of vapour pressure, level plot may be unnesc could just have std plot
\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/vap_ras.pdf} 
	\caption{Raster plot of vapour pressure 9am } 
	\label{fig:vap_ras} 
\end{figure}

Figure~\ref{fig:vap_ras} shows a level plot of the vapour pressure at 9am \citep{rastervis}. The pixel values of the vapour pressure across the state on the 7th of February 2009. We see very clearly in the north-west of the state vapour pressure at 9am is much lower relative to the south-east.

@@

\begin{table}
	\centering
	\begin{tabular}{lrrrrrrrr}
		\toprule
		       & ffdi  & temp  & df    & rh    & ws    & curing & rain  & vap \\
		\midrule
		ffdi   & 1.00  & 0.77  & 0.62  & -0.77 & 0.04  & 0.51   & -0.20 & 0.01 \\
		temp   & 0.77  & 1.00  & 0.43  & -0.70 & -0.18 & 0.46   & -0.10 & 0.39 \\
		df     & 0.62  & 0.43  & 1.00  & -0.58 & -0.09 & 0.47   & -0.37 & -0.08 \\
		rh     & -0.77 & -0.70 & -0.58 & 1.00  & 0.19  & -0.49  & 0.31  & 0.06 \\
		ws     & 0.04  & -0.18 & -0.09 & 0.19  & 1.00  & 0.01   & 0.07  & -0.16 \\
		curing & 0.51  & 0.46  & 0.47  & -0.49 & 0.01  & 1.00   & -0.11 & 0.05 \\
		rain   & -0.20 & -0.10 & -0.37 & 0.31  & 0.07  & -0.11  & 1.00  & 0.17 \\
		vap    & 0.01  & 0.39  & -0.08 & 0.06  & -0.16 & 0.05   & 0.17  & 1.00 \\
		\bottomrule
	\end{tabular}
	\caption{Correlation matrix of variables}
	\label{table:correl}
\end{table}

Table~\ref{table:correl} shows the cross correlation matrix for the variables used to predict probability of ignition. High correlation can pose a problem in regression models and to a lesser extent machine learning methods. Variables of concern are temperature, drought factor, relative humidity and FFDI and  which have relatively strong correlation coefficients.

@@

%Commented out until final printing, 
%Map of fire ignitions in Victoria
\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/data/firemap.pdf} 
	\caption{Map of fire ignition 2000 - 2009} 
	\label{fig:firemap} 
\end{figure}

%timeseries histogram with count of ignitions per fortnight
\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/data/occurence.pdf} 
	\caption{Number of ignitions by fortnight} 
	\label{fig:occ} 
\end{figure}

@@

Figure~\ref{fig:firemap} shows the locations of the fires in the dataset on the map of Victoria. The most noticeable thing in Figure~\ref{fig:firemap} is the sparsity of fires in the north-east and north west of the state and around Melbourne. This is due to the division of fire fighting organisations in Victoria. The sparse areas correspond to National Parks where Parks Victoria or the Department of Environment, Land, Water and Planning (DELWP) coordinate fire fighting activities, so many of the fires occurring in these areas are not included in this dataset.

\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/tw_fire.pdf} 
	\caption{Wind Speed and temperature coloured by fire/no fire} 
	\label{fig:tw_fire} 
\end{figure}

Figure~\ref{fig:tw_fire} is constructed in the same manner as Figure~\ref{fig:tw_fdr} but instead of grouping by the fire danger rating, I group by Fire/No Fire. Windspeed and temperature play the largest role in determining the FFDI yet there is no discernible difference between the distributions of the two variables when there is a fire compared to no fires. This is the first indicator that the FFDI alone may not be a good predictor of probability of ignition.

\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/dh_fire.pdf} 
	\caption{Drought factor and relative humidity coloured by fire/no fire} 
	\label{fig:dh_fire} 
\end{figure}

Plotting the 2D densities of the drought factor and relative humidity in Figure~\ref{fig:dh_fire}, ranges of relative humidity don't appear to vary on fire days compared to not fire days but in districts 4, 5, 6, 7, 8, 9, 10, 11, 13, 14 and 24 there is a clear difference in the densities of fire compared to no fire. Small drought factors are mostly associated with non-fire days. In districts 2, 12, 15, 16, 17, 18, 20 and 22 do not have large differences in the two densities, these districts generally have high drought factor and low relative humidity.

\begin{figure}
	\centering 
%	\includegraphics[width=1\columnwidth]{figures/Data/gv_fire.pdf} 
	\caption{Grassland curing and vapour pressure coloured by fire/no fire} 
	\label{fig:gv_fire} 
\end{figure}

In Figure~\ref{fig:gv_fire} shows the 2D densities for grassland curing and vapour pressure for the fire and non-fire days. Once again we see a clear difference in behaviour across districts.

@@

From Figure~\ref{fig:occ} it is clear the majority of bushfires occur during the fire season. I reduce the size of the dataset, taking only observations in Victorian fire season (October to March) for a number of reasons. First, from a policy perspective the FFDI, fire danger rating systems, grassland curing are only calculated during the fire season so using these variables outside the time periods they are available does not make intuitive sense. From a practical perspective, this paper is looking at predicting the types bushfires that threaten lives and property which we know to occur in the summer months.

@@

This final dataset is divided into a testing (30\%) and training set (70\%) randomly sampling across `Fire' and `No Fire' such that both datasets have the same proportion of `Fire' (approximately 0.05).

@@

As a result the logistic model becomes
\begin{equation}
\label{eq logits}
\Pr(Y=1|X)=\frac{\exp(\sum_{k=1}^K f_k(x_k))}{1+\exp(\sum_{k=1}^K f_k(x_k)}
\end{equation}
where $Y$ denotes fire occurrence, $x_k$ denotes the $k$th predictor and each $f_k$ represents a penalized regression spline.

@@

I address the class imbalance problem in two ways, first in building the model I create balanced samples by under sampling the `no fire' class in a similar manner to \citep{padilla11}.  The second approach is to assess each model's performance in predicting the test data using `balanced accuracy' rather than overall accuracy \citep{mosley13, japkowicz00}.

Under-sampling the no fire class addresses the problem of imbalanced classes but I also use it to reduce the serial and spatial correlation that is present in the dataset by using a `near-maximin' sampling method thereby ensuring all the data points in the sample are separate from each from each other in space and time to reduce the impact of correlation.

@@

%eg confusion matrix with ABCD
\begin{table}
	\centering
	\begin{tabular}{lrrrr}
    & 	\multicolumn{1}{l}{} & \multicolumn{2}{c}{\emph{Model Prediction}} \\
    & 	\multicolumn{1}{l}{} & \multicolumn{1}{r}{No Fire} & \multicolumn{1}{r}{Fire} \\
		\cline{3-5}
\emph{Actual} & No Fire & $a$   & $b$   & $a+b$ \\
                 & 		Fire & $c$   & $d$   & $c + d$ \\
		\cline{3-5}
                 & 		     & $a+c$ & $b+d$ & $a+b+c+d$ \\
		\cline{3-5}
	\end{tabular}
	\caption{Example confusion matrix}
	\label{table:ex}
\end{table}
An example confusion matrix is given in Table~\ref{table:ex}. 
From this matrix, we can compute accuracy, sensitivity (correctly classified fires dividied by total number of actual fires), specificity (correctly identified non-fires divided by total number of non-fires), balanced accuracy and kappa:
\begin{align}
\label{eq acc}
\mbox{Accuracy} &= \frac{a + d}{a+b+c+d} \\
\label{eq sens}
\mbox{Sensitivity} &= \frac{a }{a+b} \\
\label{eq spec}
\mbox{Specificity} &= \frac{d}{c+d} \\
\label{eq BA}
\mbox{Balanced Accuracy} &= \frac{\frac{a }{a+b} + \frac{d}{c+d} }{2} \\
\label{eq:K}
\kappa &= \frac{P_0 - P_c}{1-P_c}
\end{align}
where $P_0$ is the total agreement probability and $P_c$ is the Agreement probability which is due to chance (see \cite{kappa} for definitions).

@@

Logistic regression output for the first three models is given in Table~\ref{table:coefficients}.

@@

I add splines to Model 3 in stages using the confusion matrices for 10-fold cross validation to assess accuracy (using the caret package by \cite{caret}). The best performing model based on balanced accuracy and number of false negatives (incorrectly predicted `No Fires') is reported in Table~\ref{table:logf}.

@@

Figure~\ref{fig:gam_pl} shows the spline functions for each of the inputs. In logistic models the spline functions are not the marginal effects of the variables on probability of ignition but the marginal effect on the log-odds ratio. Equation~\ref{eq logits} can be rearranged to give the log-odds ratio \citep{james13},
\begin{equation}
\label{eq:logodds}
\log(\frac{p(X)}{1-p(X)}) = \sum_{k=1}^{K} f_k (x_k).
\end{equation}

The spline function fitted on the FFDI although non-linear, according to the standard error is not statistically different from zero (also seen in the p-values in Table~\ref{table:logf}). After taking into account the other variables in the model, FFDI does not appear to have an impact on the log-odds ratio. This may be because the other variables in the model take into account any information contained in the FFDI.

@@

%Output from best performing splines model (used xtable rather than texreg, too inconsistent?)
\begin{table}
	\centering
	\begin{tabular}{rrrrr}
		\toprule
		& Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
		\midrule
		(Intercept) & -3.1484 & 0.2757 & -11.4210 & 0.0000 \\ 
		ns(ffdi, 5)1 & -0.2905 & 0.3613 & -0.8043 & 0.4213 \\ 
		ns(ffdi, 5)2 & -0.2304 & 0.4625 & -0.4982 & 0.6183 \\ 
		ns(ffdi, 5)3 & 0.4040 & 0.5589 & 0.7229 & 0.4697 \\ 
		ns(ffdi, 5)4 & -0.4340 & 0.9674 & -0.4487 & 0.6537 \\ 
		ns(ffdi, 5)5 & -0.8815 & 0.7271 & -1.2124 & 0.2254 \\ 
		ns(temp, 1) & 1.8543 & 0.3416 & 5.4276 & 0.0000 \\ 
		ns(rh, 2)1 & 0.9325 & 0.5687 & 1.6397 & 0.1011 \\ 
		ns(rh, 2)2 & -0.6173 & 0.3607 & -1.7115 & 0.0870 \\ 
		ws & -0.0169 & 0.0146 & -1.1541 & 0.2485 \\ 
		ns(df, 5)1 & 1.1093 & 0.2798 & 3.9639 & 0.0001 \\ 
		ns(df, 5)2 & 1.0156 & 0.3115 & 3.2601 & 0.0011 \\ 
		ns(df, 5)3 & 0.7113 & 0.2604 & 2.7319 & 0.0063 \\ 
		ns(df, 5)4 & 0.8071 & 0.5479 & 1.4730 & 0.1408 \\ 
		ns(df, 5)5 & -0.1512 & 0.1950 & -0.7752 & 0.4382 \\ 
		ns(curing, 6)1 & 1.5648 & 0.1388 & 11.2754 & 0.0000 \\ 
		ns(curing, 6)2 & 2.3986 & 0.1874 & 12.7997 & 0.0000 \\ 
		ns(curing, 6)3 & 1.6637 & 0.1640 & 10.1436 & 0.0000 \\ 
		ns(curing, 6)4 & 1.6949 & 0.1335 & 12.6921 & 0.0000 \\ 
		ns(curing, 6)5 & 1.4634 & 0.3436 & 4.2588 & 0.0000 \\ 
		ns(curing, 6)6 & -0.4909 & 0.0817 & -6.0066 & 0.0000 \\ 
		rain & -0.0286 & 0.0056 & -5.0703 & 0.0000 \\ 
		vap & -0.0003 & 0.0067 & -0.0504 & 0.9598 \\ 
		\bottomrule
	\end{tabular}
	\caption{Regression output for Model 4}
	\label{table:logf}
\end{table}

%GAM plots, not sure if this is exactly splines... 
\begin{figure}
\null\vfill\noindent
%\includegraphics[page=1,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=2,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
%\includegraphics[page=3,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=4,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
%\includegraphics[page=5,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\hfill
%\includegraphics[page=6,width=.48\linewidth,height=.45\textheight,keepaspectratio]{figures/gam_pl.pdf}
\vfill
\null
	\caption{Plots of smoothing functions in Model 4 with 95\% point-wise confident interval} 
	\label{fig:gam_pl} 
\end{figure}

The focus of this project is to build an accurate predictor of fire ignition, so out of sample prediction is a more important evaluation tool than hypothesis testing. Table~\ref{table:cm2} contains confusion matrices (values are proportions of total number of observations in the test set) based on test data predictions, using a cutoff value of 0.5. The confusion matrix is accompanied by measures of accuracy in Table~\ref{table:acc2}.

\todo[inline]{These confusion matrices are transposed compared to Table 3.}
